Statistical Machine Learning Models on A Bank Marketing dataset

### 2025-12-08 Slides.Rmd
Fixed KNN and Boosting Trees

## KNN 1/3: Feature Preparation 
* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
```{r}
library(dplyr)
library(class)
library(caret)

@@

```
```{r}
library(class)

@@

## KNN 2/3: Hyperparameter Tuning 
* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
```{r}
library(class)
library(caret)

@@

```{r}
library(pROC)
library(caret)

@@

## Boosting 1/3: Model Formulation
* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
```{r}
library(gbm)

@@

## Boosting 2/3: Hyperparameter Tuning 
* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
```{r}
boost_model <- gbm(
  formula = y_num ~ . - y,
  data = bank_train,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  verbose = FALSE
)

@@

## Boosting 3/3: Evaluation & Importance 
* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
```{r}
library(pROC)

>>>>

## KNN 1/3: Feature Preparation 
* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
```{r, echo=FALSE, results='asis'}

library(dplyr)
library(class)
library(caret)

@@

```
```{r, echo=FALSE, results='asis'}
library(class)

@@

## KNN 2/3: Hyperparameter Tuning 
* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
```{r, echo=FALSE, results='asis'}
library(class)
library(caret)

@@

```{r, echo=FALSE, results='asis'}
library(pROC)
library(caret)

@@

## Boosting 1/3: Model Formulation
* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
```{r, echo=FALSE, results='asis'}
library(gbm)

@@

## Boosting 2/3: Hyperparameter Tuning 
* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
```{r, echo=FALSE, results='asis'}
boost_model <- gbm(
  formula = y_num ~ . - y,
  data = bank_train,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  verbose = FALSE
)

@@

## Boosting 3/3: Evaluation & Importance 
* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
```{r, echo=FALSE, results='asis'}
library(pROC)
